---
title: "Practical Machine Learning Course Project"
author: "Anthony Rosati"
date: "Sunday, November 22, 2015"
output: html_document
---

# Practical Machine Learning Course Project
========================================================

## Synopsis: 

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

The goal of this project is to predict the manner of performing unilateral dumbbell biceps curls based on data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. The 5 possible methods include -
* A: exactly according to the specification 
* B: throwing the elbows to the front
* C: lifting the dumbbell only halfway 
* D: lowering the dumbbell only halfway
* E: throwing the hips to the front

#### Load libraries and setup working directory
````{r}
rm(list = ls(all = TRUE))

library(caret)

trainingRaw <- read.csv(file="data/pml-training.csv", header=TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep=',', na.strings=c('NA','','#DIV/0!'))
testingRaw <- read.csv(file="data/pml-testing.csv", header=TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep=',', na.strings=c('NA','','#DIV/0!'))

trainingRaw$classe <- as.factor(trainingRaw$classe)  
````

#### Examine the data

#### Cleaning variables
After investigating all the variables of the sets, it's possible to see a lot of NA values, or useless or empty variables for the predictions. Since we were requested to compute the prediction only on the accelerometers values of belt, forearm, arm and dumbell, we thus discard non-accelerometer measures with the useless variables.

````{r}
NAindex <- apply(trainingRaw,2,function(x) {sum(is.na(x))}) 
trainingRaw <- trainingRaw[,which(NAindex == 0)]
NAindex <- apply(testingRaw,2,function(x) {sum(is.na(x))}) 
testingRaw <- testingRaw[,which(NAindex == 0)]
````
#### Preprocessing variables

First, we do some preprocessing of variables in the training set...
````{r}

v <- which(lapply(trainingRaw, class) %in% "numeric")

preObj <-preProcess(trainingRaw[,v],method=c('knnImpute', 'center', 'scale'))
trainLess1 <- predict(preObj, trainingRaw[,v])
trainLess1$classe <- trainingRaw$classe

testLess1 <-predict(preObj,testingRaw[,v])
````
#### Removing the non zero variables
We Remove variables with values near zero, since they have minimal impact in the predictions...

````{r}
nzv <- nearZeroVar(trainLess1,saveMetrics=TRUE)
trainLess1 <- trainLess1[,nzv$nzv==FALSE]

nzv <- nearZeroVar(testLess1,saveMetrics=TRUE)
testLess1 <- testLess1[,nzv$nzv==FALSE]
````

#### Create cross validation set
The training set is now divided in two parts, one for training and the other for cross validation...

````{r}
set.seed(98765432)

inTrain = createDataPartition(trainLess1$classe, p = 0.75, list=FALSE)
training = trainLess1[inTrain,]
crossValidation = trainLess1[-inTrain,]
````

#### Training model
Execute the training model using random forest methdology due to its high accuracy rate. The model is build on a training set of 28 variables from the initial 160. Cross validation is used as the training model's control method.

````{r}
modFit <- train(classe ~., method="rf", data=training, trControl=trainControl(method='cv'), number=5, allowParallel=TRUE )

````

#### Accuracy on training set and cross validation set
Let's examine computations on the accuracy of training and the cross validation set...

Training set:
````{r}
trainingPred <- predict(modFit, training)
confusionMatrix(trainingPred, training$classe)
````

Cross validation set:
````{r}
cvPred <- predict(modFit, crossValidation)
confusionMatrix(cvPred, crossValidation$classe)
````

#### RESULTS
Let's now run our predictions using the model on the real testing set... We will generate files for uploading and examination.

````{r}
answers <- as.character(predict(modFit, testLess1))
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("results2/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(answers)

````
Of note, all 20 predicitions from this model compared correctly against the solution set provfided by the course instructors.
